# Using Self-Organizing Maps with ```SOMbrero``` to cluster a numeric dataset

#### Laura BENDHAÏBA, Madalina OLTEANU, Nathalie VILLA-VIALANEIX

## Basic package description

To be able to run the SOM algorithm, you have to load the package called 
```SOMbrero```. The function used to run it is called ```trainSOM()``` and is 
detailed below.

__This documentation only considers the case of numerical data.__

```{r loading,results='hide',echo=FALSE,warning=FALSE, message=FALSE}
library(SOMbrero)
# source('../../R/grids.R')
# source('../../R/parameters.R')
# source('../../R/som.R')
# source('../../R/plots.R')
# library(wordcloud)
# library(igraph)
```

### Arguments

The ```trainSOM``` function has several arguments, but only the first one is
required. This argument is ```x.data``` which is the dataset used to train the 
SOM. In this documentation, it is passed to the function as a matrix or a data
frame with numeric variables in columns and observations of these variables in
rows.

The other arguments are the same than the arguments passed to the ```initSOM```
function (they are parameters defining the algorithm, see ```help(initSOM)```
for further details).

### Outputs

The ```trainSOM``` function returns an object of class ```somRes``` (see 
```help(trainSOM)``` for further details on this class).

## First case study: simulated data in $latex \mathbb{R}^2 $

The first case study show the clustering of points randomly distributed in the
square $latex [0,1]^2$. The data are generated by:
```{r dataGeneration}
set.seed(4031719)
the.data <- data.frame("x1"=runif(500), "x2"=runif(500))
plot(the.data, pch=19)
```

### Training the SOM
The numeric SOM algorithm is used to cluster the data:
```{r dataTrain}
set.seed(4031731)
# run the SOM algorithm with 10 intermediate backups and 2000 iterations
my.som <- trainSOM(x.data=the.data, dimension=c(5,5), nb.save=10, maxit=2000, 
                   scaling="none")
```

The energy evolves as described in the following graphic:
```{r energy}
plot(my.som, what="energy")
```
It is stabilized during the last 500 iterations.

### Clustering

The resulting clustering distribution can be visualized by the hitmap:
```{r hitmapObs}
plot(my.som, what="obs", type="hitmap")
```
The observations are almost uniformly distributed on the map.

The clustering component allows us to plot the initial data according to the 
final clustering.

```{r clusteredData, echo=FALSE, cache=TRUE}
# prepare a vector of colors
my.colors <- rainbow(prod(my.som$parameters$the.grid$dim))[my.som$clustering]

# points depicted with the same color are in the same final cluster
plot(my.som$data[,1], my.som$data[,2], col=my.colors, pch=19, xlab="x1", 
     ylab="x2", main="Data according to final clustering")
```

### Clustering interpretation

The values of the prototypes can be represented with the plot function and help
interpret the clusters:
```{r colorProto, fig.width=5, fig.height=2.5}
par(mfrow=c(1,2))
plot(my.som, what="prototypes", type="color", var=1, main="prototypes - x1")
plot(my.som, what="prototypes", type="color", var=2, main="prototypes - x2")
```
Here, the interpretation is simple enough: high values of the first variables x1
are located at the bottom of the map and small values at the top of the map.
Large values of x2 are located at the right hand side of the map, whereas, small
values are located at the left hand side of the map.

We obtain the same results with the same kind of graphic on the observation mean 
values:
```{r colorObs, fig.width=5, fig.height=2.5}
par(mfrow=c(1,2))
plot(my.som, what="obs", type="color", var=1, main="obs mean values - x1")
plot(my.som, what="obs", type="color", var=2, main="obs mean values - x2")
```

The prototypes coordinates are also registered for each intermediate backup so 
they can be displayed on different graphics to see the evolution in the
prototypes organization.
```{r protoEvoluation, fig.width=15, fig.height=6, echo=FALSE}
# find out the prototypes to be linked
tra <- NULL
for (i in c(1,6,11,16,21)){
  tra <- c(tra, i, i+1, i+1, i+2, i+2, i+3, i+3, i+4)
}
for (i in c(1:5)){
  tra <- c(tra, i, i+5, i+5, i+10, i+10, i+15, i+15, i+20)
}
tmp <- matrix(tra, ncol=2, byrow=TRUE)
# plot the prototypes
par(mfrow=c(2, 5),mar=c(3,2,2,1))
invisible(sapply(1:my.som$parameters$nb.save, function(ind){
  plot(my.som$backup$prototypes[[ind]][,1], my.som$backup$prototypes[[ind]][,2],
       xlab="", ylab="", main=c("iteration ", my.som$backup$steps[ind]))
  for (i in 1:nrow(tmp)){
    segments(x0=my.som$backup$prototypes[[ind]][tmp[i,1],1], 
             y0=my.som$backup$prototypes[[ind]][tmp[i,1],2],
             x1=my.som$backup$prototypes[[ind]][tmp[i,2],1], 
             y1=my.som$backup$prototypes[[ind]][tmp[i,2],2], 
             col="red", pch=19)
  }
}))
```
At the begining of the algorithm, the prototypes are randomly distributed in
$latex \mathbb{R}^2$ and then, they organize as a regular rectangular grid in
$latex [0,1]^2$.

## Second case study: the iris dataset

This second case study is performed on the famous (Fisher's or Anderson's) iris
data set that gives the measurements in centimeters of the variables sepal
length and width and petal length and width, respectively, for 50 flowers from
each of 3 species of iris (setosa, versicolor, and virginica).

### Training the SOM

The first four variables of the data set (that are the numeric variables) are 
used to map the each flower on the SOM grid.
```{r irisTrain, cache=TRUE}
set.seed(4031730)
# the following instruction run the SOM algorithm with verbose set to TRUE
iris.som <- trainSOM(x.data=iris[,1:4],verbose=TRUE, nb.save=5)
iris.som
```

As the energy is registered during the intermediate backups, we can have a look
at its evolution.
```{r energyIris}
plot(iris.som, what="energy")
```
which is stabilized during the last 100 iterations.


### Resulting clustering

The clustering component contains the final classification of the dataset. It is 
a vector having a length equal to the number of rows of the input dataset.
```{r irisClusters}
iris.som$clustering
table(iris.som$clustering)
```
which can also be visualized by a hitmap plot:
```{r irisHitmap}
plot(iris.som, what="obs", type="hitmap")
```

To access the relevance of each explanatory variable in the definition of the 
clusters, the function ```summary``` includes an ANOVA with the factor being the
clustering, for each (numeric) input variables.
```{r irisSummary}
summary(iris.som)
```
Here, all variables have significantly different means among the different
clusters and can thus be considered to be relevant for the clustering
definition.

Another usefull function is ```predict.somRes```. This function predicts the 
neuron to which a new observation would be assigned. The first argument must be 
a ```somRes``` object and the second one the new observation.
Let us have a try on the first observation of the iris data set:
```{r irisPred1}
# call predict.somRes
predict.somRes(iris.som, iris[1,1:4])
# check the result of the final clustering with the SOM algorithm
iris.som$clustering[1]
```


### Clustering interpretation

#### Common graphics to observations and prototypes

__NB: in all the following graphics, data are scaled in the preprocessing
stage.__

Some graphics are shared between observations and prototypes. They allow you 
to plot the level values, either of the prototypes or of the observation means. 
In this example these commun graphics are plotted for the mean observation 
values.

```{r irisGraphOP}
par(mfrow=c(2,2))
plot(iris.som, what="obs", type="color", variable=1, print.title=TRUE, 
     main="Sepal length")
plot(iris.som, what="obs", type="color", variable=2, print.title=TRUE, 
     main="Sepal width")
plot(iris.som, what="obs", type="color", variable=3, print.title=TRUE, 
     main="Petal length")
plot(iris.som, what="obs", type="color", variable=4, print.title=TRUE, 
     main="Petal width")
plot(iris.som, what="prototypes", type="lines", print.title=TRUE)
plot(iris.som, what="obs", type="barplot", print.title=TRUE)
plot(iris.som, what="obs", type="radar", print.title=TRUE)
```

The first thing that catch the attention is these 2 empty squares: none of the 
observations is assigned neither to cluster 14 nor to cluster 19. 

Let us analyse more precisely the obtained results for cluster 5.
On the ```"color"``` plots, the cluster 5 has the following results: middle 
level value for ```Sepal.Length```, high value for ```Sepal.Width``` and low 
values for ```Petal.Length``` and ```Petal.Width```.
These results are also noticeable in the other plots:

* on the ```"lines"``` plot, the first point is quite in the middle, the second 
one is very high and the 2 other are low;
* on the ```"barplot"``` plot, you can observe again this evolution: first a 
peak and a fall to the minimum value;
* on the ```"radar"``` plot, the slice corresponding to ```Sepal.Width``` is 
very big whereas the 2 Petal slices are almost invisible.

Flowers with large and long petals and a long sepal are classified in the left
bottom corner of the map. Flowers with a large sepal and a small petal (either
length and width) are classified in the left upper corner of the map.

#### More graphics on observations

```{r irisObs}
plot(iris.som, what="obs", type="boxplot", print.title=TRUE)
rownames(iris)
plot(iris.som, what="obs", type="names", print.title=TRUE)
```

* By default, the ```"boxplot"``` plot considers all numerical variables of the 
data set. As we used in this example the 4 numerical variables of the iris data 
set, 4 boxplots are plotted in each neuron.
* The ```"names"``` plot prints, for one neuron, the names of the observations 
which are assigned to this cluster. Here, the row names of iris data set are 
simply the number of the lines.

#### More graphics on prototypes

Some more graphics handling prototypes have been implemented.

```{r irisProto}
par(mfrow=c(2,2))
plot(iris.som, what="prototypes", type="3d", variable=1, main="Sepal length")
plot(iris.som, what="prototypes", type="3d", variable=2, main="Sepal width")
plot(iris.som, what="prototypes", type="3d", variable=3, main="Petal length")
plot(iris.som, what="prototypes", type="3d", variable=4, main="Petal width")
```
```"3d"``` provides the same results as "color" but in a 3 dimensional graphic:
x is the x dimension of the grid, y is the y dimension of the grid and z is the
value of the prototype for the variable ```variable``` (by name or number in the
dataset) of the corresponding neuron.

Also, some graphics are provided to visualize the distance between prototypes on
the grid:
```{r irisDistProto}
plot(iris.som, what="prototypes", type="poly.dist")
plot(iris.som, what="prototypes", type="umatrix")
plot(iris.som, what="prototypes", type="smooth.dist")
plot(iris.som, what="prototypes", type="mds")
plot(iris.som, what="prototypes", type="grid.dist")
```
* ```"poly.dist"``` plots, for each neuron, a polygon that has vertex 
coordinates representing the distance matrix between prototypes. The 
colors indicates the number of observations in the neuron;
* ```"umatrix"``` fills the neurons of the grid using colors that represent
the average distance between the current prototype and its neighbors.
* ```"smooth.dist"``` plots the mean distance between the current prototype and 
its neighbors with a color gradation.
* ```mds``` plots the number of the neuron on a map according to a Multi
Dimensional Scaling (MDS) projection.
* ```grid.dist``` plots points which x coordinates are the distances between 
all prototypes and which y coordinates are the distances between all neurons of 
the grid.


#### Graphics according to an additional variable

##### To a factor

Let us plot pies for all neurons according to the flower species of the 
observations.
```{r irisAdd1}
class(iris$Species)
levels(iris$Species)
plot(iris.som, what="add", type="pie", variable=iris$Species)
```

##### To a numerical vector

The ```"color"``` plot available for ```"add"``` is similar to the ```"obs"``` 
or ```"prototypes"``` cases. Here we choose as an additional variable the first 
variable of the iris data set, so we obtain the same graphic as above (cf. 
section __Common graphics to observations and prototypes__).
```{r irisAdd2}
plot(iris.som, what="add", type="color", variable=iris$Sepal.Length)
```

##### To a numerical matrix or data frame

The ```"lines"```, ```"barplot"```, ```"radar"``` and ```"boxplot"``` plots 
available for ```"add"``` are similar to the ```"obs"``` or ```"prototypes"``` 
cases.

```"words"``` is only implemented for an additional variable. In this case, the 
additional variable must be a contingency matrix: the words used on the graph 
are the names of the colums and the presence or lack of the word is expressed 
by respectively 1 or 0. The size of the words on the grid depends on the rate 
of presence in the observations of the current neuron.
```{r irisMatCont, echo=FALSE}
my.cont.mat <- matrix(data=c(rep(c(rep(1,50), rep(0,150)), 2), rep(1,50)), 
                      nrow=150, ncol=3)
colnames(my.cont.mat) <- levels(iris$Species)
```
```{r irisAdd4}
# my.cont.mat is the contingency matrix corresponding to the variable 
# iris$Species - overview of the 5 first lines:
my.cont.mat[1:5,]
plot(iris.som, what="add", type="words", variable=my.cont.mat)
```

##### To a non numerical vector

```"names"``` is similar to the ```"names"``` case implemented for ```"obs"```. 
Here we choose to give to the argument ```variable``` the row names of the iris 
data set: so we obtain the same graphic as above (cf. 
__More graphics on observations__).
```{r irisAdd5}
plot(iris.som, what="add", type="names", variable=rownames(iris))
```

Then, we can try to call this plot again but on the variable ```iris$Species```.
```{r irisAdd5bis}
plot(iris.som, what="add", type="names", variable=iris$Species)
```
We obtain exactly the same plot as we obtained above for ```"words"``` with the 
contingency matrix corresponding to the variable ```iris$Species```.

## Third case study: data from package ```GeneNet```

Our last case study is performed on the data ```arth800``` from the
package ```GeneNet```.

### Train the SOM algorithm

As it is figured out by the ```summary``` function, this data set provides 
measurements on 22 variables (22 genes) at 11 different time points. As there 
are some repeated measurements, there are 800 measurements per variable (per 
gene).

```{r GeneNetSOM}
data(arth800, package="SOMbrero")
summary(arth800.expr)
```

```{r trainARTH800, cache=TRUE}
set.seed(1632)
gene.som <- trainSOM(x.data=arth800.expr, dimension=c(5,5))
```

### Clustering interpretation

##### To an ```igraph``` object

The last available plot for a ```somRes``` object is ```"graph"```. Then, the 
argument ```variable``` must be given an ```igraph``` object. A new graph is 
plotted where a vertex represents a neuron and the width of an edge between 2 
vertices is proportional to the number of edges in the ```igraph``` object 
between the observations of these 2 neurons.
Here we chose the graph inferred from the ```arth800``` expression data using
the functions ```ggm.estimate.pcor``` and ```ggm.test.edges``` of the package
```GeneNet```. This graph is included in the example dataset in the **R**
```igraph``` object ```arth800.graph```.
```{r GeneNetGraph}
class(arth800.graph)
summary(arth800.graph)
plot(gene.som, what="add", type="graph", pie.graph=FALSE, 
     variable=arth800.graph, edge.curved=TRUE, s.radius=0.3, print.title=TRUE)
```
The graph vertices are the neurons of the grid. If there is a vertex missing, 
it means that this vertex is empty of observations. The width of the edge 
between 2 vertices is proportional to the number of edges of the original graph
between the observations classified in the corresponding neurons.
